---
output:
  html_document: default
  pdf_document: default
---
# Red Wine EDA
Tiago B.N. Silveira, MSc.  
July 15, 2018  

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(ggpubr)
library(GGally)
```

```{r echo=FALSE,}
# Load the Data
wines <- read.csv("wineQualityReds.csv")
```

This project is part of the Udacity Data Analysis Nanodegree and aims to apply 
**exploratory data analysis (EDA)** techniques using R in order to find out
relationships in one to multiple variables, as well as exploring a given dataset
for distributions, outliers, and anomalies. 

EDA can thus be understood as the numerical and graphical examination of data 
characteristics and relationships before formal, rigorous statistical analyses 
are applied. EDA can lead to insights, which may uncover to other questions, 
and eventually predictive models. As explained during the Udacity course, EDA 
is also an important “line of defense” against bad data and is an opportunity 
to notice that some assumptions or intuitions about a dataset are violated.

## 1. Understanding the data
The first step before diving down EDA is to provide a short introduction into 
the dataset we are planning to explore. In this case, from a list of available
options I have chosen the "Wine Quality" dataset ([available at http://dx.doi.org/10.1016/j.dss.2009.05.016](http://dx.doi.org/10.1016/j.dss.2009.05.016)) 
since I'm a wine lover. 

Said that, and since our data is already loaded, the first thing to do is to 
take a look at the very first line of the dataset:

```{r echo=FALSE, Load_the_Data}
# First look at dataset
head(wines)
str(wines)
summary(wines)
```

We already knew it is a tidy dataset. Anyway, it was possible to check there is
any missing value (NA values, for example) through the *summary* output.

To understand each data attributes, it's helpful to consult the data description
where the following information is found:  

> The two datasets are related to red and white variants of the Portuguese 
"Vinho Verde" wine. For more details, consult: http://www.vinhoverde.pt/en/ or 
the reference [[1]](#[1]). Due to privacy and logistic issues, only 
physicochemical (inputs) and sensory (the output) variables are available (e.g. 
there is no data about grape types, wine brand, wine selling price, etc).  

It's also important to notice that besides the target feature is the wine 
quality, the dataset is unbalanced regarding to it. It means there are few 
excellent or poor wines. This issue will be addressed later. For now, let's 
start taking a look at each attribute in order to understand our data in deepen
way.


## 2. Univariate Analysis
In order to catch the meaning of each variable in the dataset, in this section 
we will seek to explore each attribute in what is known as *univariate 
analysis*. The most relevant attributes of this dataset are consolidated at the
end of this section.  


### A closer look at each attribute
#### Quality
Since it's the target attribute, there is any doubt about its relevance. That
way, let us first understand its distribution through a bar chart, since quality
is an ordinal data.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.quality}
ggplot(aes(x=quality), data = wines) + 
    geom_bar(fill = 'steelblue', colour='darkgrey', alpha= 0.8)
```

The first thing to notice is that there are few *really bad wines*, i.e. those
whose quality is below 3. In the same way, there are few *really good wines*, 
ie, those whose quality is above 8. It's also effort to notice this distribution
seems like a normal one.  

Considering this is the target attribute, all the upcoming bivariate analysis
will be in relation to this attribute, in the next section. By now, let's take a
look on the distribution of the other dataset variables, together with a boxplot
in order to show out the outliers. 

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.function}
wine_attribute <- function(attribute, varName = '') {
  ## Building the histogram:
  histogram <- ggplot(data = wines) +
      geom_histogram(aes(x=attribute), 
                     fill = 'steelblue', colour='darkgrey', alpha= 0.8) +
      labs(x = varName)
  ## Histogram scaling y_log10:
  histYlog <- histogram + scale_y_log10() +
      labs(y = 'log10(count)', x= varName)
  
  ## Histogram scaling x_log10:
  histXlog <- histogram + scale_x_log10() + 
      labs(x = paste('log10(', varName,')'))

  ## Building the boxplot highlighting the outliers:
  outliers <- ggplot(wines, aes(x = 1, y = attribute)) + 
      geom_jitter(alpha = 0.1 ) +
      geom_boxplot(alpha = 0.2, color = 'red') + 
      labs(x ='distance from mean', y= varName)
  
  ## Arranging all the plots:
  histPlots <- ggarrange(histogram, histXlog, histYlog, ncol=1, nrow=3)
  ggarrange(outliers, histPlots,ncol=2, widths = c(1,1.5))
}
```


#### Alcohol
This attribute refers to the percent alcohol content of the wine (% of volume).

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.alcohol}
## How the "alcohol" attribute is distributed?
wine_attribute(wines$alcohol, varName = 'Alcohol (% of vol)')
```

We can notice from the boxplot there are few outliers in this attribute 
distribution, i.e. most of the samples stands inside $1.5$ times the 
interquartile rate (IQR). From the histogram, the cartesian coordinates are 
adequate to understand this variable distribution, which is concentrated on 9% 
values.  


#### Sulphates
Accordingly to the dataset documentation, it refers to a wine additive which 
can contribute to sulfur dioxide gas ($S0_2$) levels. This last acts as an 
antimicrobial and antioxidant. The sulphates levels are measured by the 
concentration of *potassium sulphate* in *g/dm³*.  

To find out the range of sulphates in the evaluated wines, we can just summarize
this attribute, as follow:

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.sulphate.summary}
summary(wines$sulphates)
```

As we did before, the next graphs show how this attribute is distributed along 
our data:

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.sulphate.charts}
## How the "sulphate" attribute is distributed?
wine_attribute(wines$sulphates, varName = 'Sulphates (g/dm³)')
```

The values of this attribute are distributed around the mean showing also some
skewness, which is justified by the high number of outliers, as shown in the
boxplot.  

#### pH
This attribute describes how acidic or basic a wine is on a scale from 0 (very 
acidic) to 14 (very basic); most wines are between 3-4 on the pH scale.  

Again, our first step will be checking the values for this attribute using the
*summary* command, whose values around 3.31 confirm the above statement. 

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.ph.summary}
summary(wines$pH)
```

Trying to see this same information through the histogram, in order to confirm 
visually if it follows a normal distribution:  

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.ph.hist}
## How the "pH" attribute is distributed?
wine_attribute(wines$pH, varName = 'pH')
```

As shown above by the plots, the wine pH follows a quasi-normal distribution. 
We can also confirm that it varies in the range of 3.0 to 3.6, also with some 
outliers beyond this range.  


#### Density
The wine densities vary accordingly to density of the water and on the percent
alcohol and sugar content. I particularly prefer full-bodied wines to those 
lighter ones, characteristics I immediately associate to density. But as far as
I know, there are both strong and soft great wines, in the way I don't think 
density would be a decision attribute for wine quality. I will check this 
assumptions in the bivariate analysis. For now, let's see how this attribute
is distributed in our data:

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.density.summary}
summary(wines$density)
```

From the summary information, the variance is about 0.01 or 1%. It means a
variation of just *1g* for each *100ml* of wine. I wonder if humans are able to
perceive such small variance, which has lead me to an interesting reference 
[[2]](#[2]) where it is said that the term "body" or "weight" of a wine, besides
is almost always a desirable aspect, is a wine property not fully defined. In
the same reference, what called my attention is that *sweetness* and the 
presence of *glycerol* can increase the perception of body, while acidity 
appears to reduce the perception of body. This description has clarified my 
confusion around *wine density*, in the way we now follow with the analysis:

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.density.hist}
## How the "density" attribute is distributed?
wine_attribute(wines$density, varName = 'Density (g/cm³)')
```


#### Sulfur dioxide
As explained in the dataset description, the *sulfur dioxide* $SO_2$ can be
found in its free-form, as a dissolved gas or as a bisulfite ion, or in bounded
forms with other molecules. Besides its function to prevent microbial growth and
wine oxidation, it is mostly undetectable in wine. Free concentrations over *50 
ppm*, though, turns it evident in the nose and taste of wine.  

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.sulfur.summary}
summary(wines[c(7,8)])
```

The summary above shows out there must be an outlier on data. 

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.sulfur.boxplot}
## Reference for a univariate boxplot using ggplot2:
## https://stackoverflow.com/questions/25049104/why-does-a-boxplot-in-ggplot-requires-axis-x-and-y
wine_attribute(wines$free.sulfur.dioxide, 
               varName = 'Free Sulfur Dioxide (mg/dm³)')

wine_attribute(wines$total.sulfur.dioxide, 
               varName = 'Total Sulfur Dioxide (mg/dm³)')
```

From the boxplot it's clear to see there are two outliers in the total 
measurements whose amount of sulfur dioxide goes above $200 {mg}/{dm}^3$. I will
back to this point later during the multivariate analysis. 

In the charts above I intended to plot the histogram both for the free form and
the total amount of sulfur dioxide. It is clear to observe that the total amount
follows the same shape as for the free form, the first showing a higher 
concentration as expected.  


#### Chlorides
This attribute stands for the amount of salt (*sodium chloride*) in the wine, 
measured in $g/{dm}^3$, i.e. in grams per liters. 

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.chlorides.charts}
## How the "chlorides" are present in the wines?
wine_attribute(wines$chlorides, varName = 'Chlorides (g/dm³)')
```

From the boxplot we notice there are outliers in the right-skewed chlorides 
distribution. From the histogram, these outliers are visible when we use a log10
scale for y-axis.  


#### Residual sugar
This attribute stand for the amount of sugar remaining after fermentation stops. 
Accordingly to the dataset documentation, it's rare to find wines with less than
1 gram/liter (or $g/{dm}^3$ in the International System of Units - SI). On the
other hand, wines with more than 45 grams/liter of sugar are categorized as 
sweet.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.sugar}
wine_attribute(wines$residual.sugar, varName = 'Residual sugar (g/dm³)')
```

Analyzing the histogram above, even though there are some outliers, none of them
is above $45 g/{dm}^3$, which means that the whole dataset is about dry wines.


#### Acidity
There are three attributes in the dataset related to the wine acidity, all of
them measured in $g/{dm}^3$:  
* **Fixed acidity**: normally refers to *tartaric acid* which are involved with 
wine and do not evaporate readily.  
* **Volatile acidity**: the amount of acetic acid in wine, which at too high 
levels can be unpleasant, tasting like a vinegar.  
* **Citric acid**: usually found in small quantities, citric acid can add some 
*freshness* and flavor to wines. 

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate.acidity.summary}
## For this attribute I will plot in a different layout than the previous ones:
vertPlot <- function(attribute, varName = '') {
  ## Building the boxplot highlighting the outliers:
  outliers <- ggplot(wines, aes(x = 1, y = attribute)) + 
      geom_jitter(alpha = 0.1 ) +
      geom_boxplot(alpha = 0.2, color = 'red') + 
      labs(x ='distance from mean', y= varName)
  ## Building the histogram:
  histogram <- ggplot(data = wines) +
      geom_histogram(aes(x=attribute), 
                     fill = 'steelblue', colour='darkgrey', alpha= 0.8) +
      labs(x = varName)
  ## Histogram scaling x_log10:
  histXlog <- histogram + scale_x_log10() + 
      labs(x = paste('log10(', varName,')'))
  
  ## Arranging the plots vertically:
  ggarrange(outliers, histogram, histXlog, ncol=1, nrow=3, 
            heights = c(0.5,0.25,0.25))
}

## Creating the plots for each type of acid:
a1 <- vertPlot(wines$citric.acid, varName = 'Citric acid (g/dm³)')
a2 <- vertPlot(wines$volatile.acidity, varName = 'Volatile acidity (g/dm³)')
a3 <- vertPlot(wines$fixed.acidity, varName = 'Fixed acidity (g/dm³)')

ggarrange(a1,a2,a3, ncol=3)

```

From these charts, and considering the description for each kind of acidity, we 
can draw as hypothesis that the *citric acid* outliers relates to some wines 
whose main characteristic is its freshness or its fruit flavor. The fact that 
both *citric acid* and *volatile acidity* log distribution have the same shape
called my attention, at first. However, as we shall see in the further analysis,
it does not mean any relationship among these attributes by now. The fixed 
acidity, on the other hand, presents values in a higher scale as well as more 
outliers.  

### Findings from univariate analysis
In this Section I explored the *Red Wine dataset* built from a scientific 
research about whose objective was to model wine preferences by data mining from
physicochemical properties [[1]](#[1]).  

This first analysis, which is called univariate since it considers only one 
attribute at a time, is essential because it allows both a general understanding
of the subject to be explored and the meaning of each available attribute in 
the dataset.

At this point, we are already able to select some important attributes in order
to check how they relate to each other:  
  
  * The **quality** attribute will be our *targeted variable* throughout the 
  whole analysis. As a premise, this analysis objective is to find out which
  attributes are related to wines of better quality, ie, those assigned with 
  a quality level $\geq 7$.  
  
  * The description obtained for each attribute, such as **alcohol**, **pH**, 
  and **density**, for example, will allow us to analyze them jointly with the
  **quality** attribute in the next section, in order to seek for correlation. 
  
  * Many attributes have presented outliers in some level. We could define a 
  strategy of outlier removal or even though a deeper analysis on them, trying
  to understand why these values are distant from the mean distribution. In this
  analysis, though, we will only describe them.
  

## 3. Bivariate Analysis  
In this section we will start exploring the relationships among all variables in
our dataset, directing the analysis to find out which attributes are mainly 
related to our target variable: the quality score. We start now by calculating 
the correlation among all the attributes: 

```{r echo=FALSE, bivariate.correlation}
## Reference: https://briatte.github.io/ggcorr/#controlling-the-main-geometry
ggcorr(wines[,2:13], geom = "blank", label = TRUE, 
       hjust = 0.9, layout.exp = 2) +
  geom_point(size = 8, aes(color = coefficient > 0, 
                           alpha = abs(coefficient) > 0.35)) +
  scale_alpha_manual(values = c("TRUE" = 0.25, "FALSE" = 0)) +
  guides(color = FALSE, alpha = FALSE)
```

From the correlation matrix we notice the only relevant attributes to quality 
are (i) the percentage of alcohol and (ii) the volatile acidity. There are some 
relevant correlation among other variables as well, and even though they don't 
contribute explicitly to the target variable we are interested in, it will be 
insteresting to take a look on them and find out some new information or 
insights.  

In order to select which of them we will work with, we should analyze the matrix 
above considering the previous univariate analysis for each attribute. In this 
case, we can draw the following assumptions:  

* The wine **quality** is positively correlated to **alcohol** and negatively 
correlated to the **volatile acidity**. It does make sence with our previous
findings, since the volatile acidity is responsible for an unpleasent taste in 
the wine.  

* It is quite obvious that the pH attribute is negatively correlated to acidity,
since the lower values in the pH scale means the higher acidity. Since it is a 
well-known fact, there is no sense to explore these relationships here.  

* There seems to be some correlation between sulphates and chlorides. However, 
none of them presents a relevant correlation with quality. Furthermore, from the
foregoing description it is possible to infer that both attributes are derived 
from the same mineral, so that we will not consider these attributes in the next
analysis.  

* It is interesting to observe there are some relationship between **density** 
and other attributes such as **alcohol**, **residual sugar**, **citric acid**, 
and **fixed acidity**.  

* It also seems to be evident that the free and total forms of sulfur dioxide
will be related to each other. The same stands for the different acidity types. 
Regarding this last, in order to simplify this analysis, we will now consider 
only the **volatile acidity**  due to its impact on quality levels.  

Based on this, our next task is to create some pairplots relating all these 
chosen attributes: 

```{r echo=FALSE, bivariate.overview}
theme_set(theme_minimal(10))
## Defining the variables we will explore:
variables <- c('quality', 'alcohol','volatile.acidity',
               'density', 'residual.sugar')

## Plotting ggpairs for the selected attributed:
ggpairs(wines[variables], aes(alpha=0.3))

## I experimented adding jitter, but the result was not satisfactory:
#ggpairs(wines[variables], aes(alpha=0.3), 
#   lower = list(continuous=wrap("points", position="jitter")))
```

These pairplots above summarize most of the information we have gotten until 
now, as for example the unbalanced distribution of quality -- ie, there are more
wines scored as 5 or 6 than those below and above these values. 


### Correlations to quality
Accordingly to the correlation matrix, *alcohol* and *volatile acidity* are the
attributes with the most relevant relationship to *quality*. My first idea was 
to create a scatter plot with a smooth line to analyze it deeply. However, this 
type of visualization is not appropriate when one of the features on the chart 
is categorical or discrete, as is the case of *quality*. A boxplot, though, 
brings so much more information, as follow:  

```{r echo=FALSE, bivariate.alcohol-quality.boxplot}
ggplot(aes(x= factor(quality), y= alcohol), data = wines) +
  geom_jitter( alpha = .3) +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", geom = "point", color = "red", 
               shape = 16, size = 4) +
  labs(x= 'Quality (score between 3 and 9)',
       y= 'Alcohol (% by volume)',
       title= 'Alcohol Vs. Quality')
```

As it was expected to, there is some correlation between the alcohol percentage 
in the wine and its quality. However, as it's clear to see from the scattered 
points, different quality levels are assigned when the wine has 11% of alcohol. 
This leads us to afirm that relevant correlation between alcohol and quality
arises only when we are dealing with the bordering values.

Another way to check it out is building a radar chart with the alcohol 
statistics for each one of the quality levels:

```{r bivariate.alcohol-quality.radar, echo=FALSE, message=FALSE, warning=FALSE}
## First, we need to select the appropriate data:
alcohol_quality <- wines %>%
  group_by(quality) %>%
  summarise(alcohol_mean = mean(alcohol))

## Since ggplot2 does not contain a ready function for radar chart, I chose to
## use the radarchart function from "fmsb" package:
#install.packages('fmsb')
library(fmsb)

## In order to apply my data to radarchart, I need to transpose it before:
alcohol_quality <- as.data.frame(t(alcohol_quality))
# The code above set the dataframe name accordingly to the first dataframe row:
names(alcohol_quality) <- paste('Quality',alcohol_quality[1,], sep = '-')
# We only want the second line, ie, we need to drop the first row:
alcohol_quality <- alcohol_quality[2,]
# In order to use the radarchart, we need to add the minmax values as rows:
alcohol_quality=rbind(rep(12.4,6) , rep(9.6,6) , alcohol_quality)

## Now plotting the radarchart:
radarchart(alcohol_quality, axistype=1, calcex = 0.8, vlcex=0.9,
           caxislabels = c('9.6%','','11%','','12.4%'),
           title="Percentage of alcohol by quality level")
```

From the chart above it becomes really clear that the wines assigned with a 
better quality in average have higher alcohol percentages.  

The next chart tries 

```{r echo=FALSE, bivariate.volatile-quality.boxplot}
ggplot(aes(x= factor(quality), y= volatile.acidity), data = wines) +
  geom_jitter( alpha = .3) +
  geom_boxplot( alpha = .5,color = 'purple')+
  stat_summary(fun.y = "mean", geom = "point", color = "red", 
               shape = 16, size = 4) +
  labs(x= 'Quality (score between 3 and 9)',
       y= 'Volatile acidity (g/dm³)',
       title= 'Volatile acidity Vs. Quality')
```

From the boxplots related to the quality attribute we can notice there are three
main categories of quality we could define: low, equivalent to the scores 3 and
4, medium, equivalent to the scores 5 and 6, and high, equivalent to the socres
7 and 8. We also can notice that the volatile acidity is negatively correlated 
to quality. 


### Other correlations
Following I try to explore other relationships among our selected variables. The
results are quite similar to those found on the pairplots, besides now with a 
better resolution and a smooth line that allows us to comprehend how this 
correlation would be for different numerical values. 

#### Alcohol Vs. Density
```{r bivariate.ph-density, echo=FALSE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
ggplot(aes(y = alcohol, x = density), data = wines) + 
  geom_point(alpha = 0.35, pch = 1) + 
  geom_smooth(method = loess) + 
  labs(y= 'Alcohol (% by volume)',
       x= 'Density (mg/dm³)',
       title= 'Relationship of Alcohol Vs. Density')
```

#### Residual sugar Vs. Density
```{r bivariate.ph-density, echo=FALSE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
ggplot(aes(y = residual.sugar, x = density), data = wines) + 
  geom_point(alpha = 0.4, pch = 1) + 
  geom_smooth(method = loess) + 
  labs(y= 'Residual sugar (g/dm³)',
       x= 'Density (mg/dm³)',
       title= 'Relationship of Residual sugar Vs. Density')
```


## 4. Multivariate Analysis
Most of the information we could gather from the dataset variables were already 
obtained in the previous section. From the previous analysis, **alcohol** is 
the attribute in our data most correlated to our target variable. In a lower 
rate, the *volatile.acidity* is also correlated. The next charts will try to 
explore the relationships among these three variables:  

```{r echo=FALSE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = volatile.acidity, color = factor(quality)), 
       data = wines) +
      geom_point(alpha = 0.8, size = 1.2) +
      geom_smooth(method = "lm", se = FALSE,size=1.2)  +
  scale_color_brewer(type='div', guide=guide_legend(title='Quality'))
```

Previously we have found that the alcohol percentage correlation to quality 
levels is higher for bordering values. From the chart above we perceive it also
becomes true when we correlate alcohol and the volatile acidity. Our next chart 
shows the boxplot for the correlation between the volatile acidity and different
buckets of alcohol percentage, accordingly to each quality level. 

```{r echo=FALSE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
wines$alcohol.bucket <- cut(wines$alcohol,c(8, 9, 10.5, 12, 16))

ggplot(aes(x = factor(quality), y = volatile.acidity), data = wines) + 
  geom_boxplot( aes(fill= wines$alcohol.bucket)) +
  scale_fill_brewer(type='seq',
                   guide=guide_legend(title='alcohol.bucket'))
```



### Linear regression models:
In this section I intend to build a linear regression model that could allow us 
to estimate the wine quality from the two most important attributes we have
found in our EDA.

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Reference: https://github.com/nicolasfguillaume/Exploratory-Data-Analysis-with-R/blob/master/Lesson_6.md
library(scales)
library(memisc)
## use the I() operator (='as is') which inhibits interpretation/conversion of 
## objects. By protecting the expression, ensures it is interpreted in the 
## obvious arithmetical way.
m1 <- lm(I(quality) ~ I(alcohol), data = wines)
m2 <- update(m1, ~ . + volatile.acidity)
mtable(m1, m2)
```

The resultant model was not as good as I expected to, with a final R-square 
measure of 32%. Since the objective of this project was limited on EDA, I will 
not try to engineer any feature in order to reach better results, but instead I
will discuss some aspects that may explain the model insuccess: 

* **Data is unbalanced**: there are few samples of the higher and lower quality
levels. Although it's not a determinant factor in linear regression, in this 
case it may have impacted.  

* **Weak correlation**: the correlation among the attributes were weak, ie. none
of them was greater than 50%.  

* **Dependent attributes**: for a better modelling, it would be better to target
exactly what we were looking for: good wines, in the case. In this analysis we
have considered the whole spectrum for quality, with the exception of some 
analyzes in which we sub-sampled the set for quality levels greater than or 
equal to 7, but whose findings were not relevant.  

* **Subjectivity**: this is the most interesting aspect to me. Unlike a diamond 
exploratory data analysis, for example, whose measures are limited to 
physicochemical factors, the main attribute of our dataset (quality) is totally
subjective. In [[3]](#[3]), for example, the influence of the human subjectivity
on wine tasting is criticized. In addition, in our dataset it was not informed
how many distinct professionals evaluated the same wine, nor the divergences 
they may have had, as well as other important aspects that should be considered.


------

# 5. Final Plots and Summary  
### Plot One
```{r echo=FALSE, Plot_One}
## Reference on how to add the title:
## http://www.sthda.com/english/rpkgs/ggpubr/reference/annotate_figure.html
## It is not used here anymore, but I will kept the references.
ggplot(aes(x = factor(quality), y = volatile.acidity), data = wines) + 
  geom_boxplot( aes(fill= wines$alcohol.bucket)) +
  scale_fill_brewer(type='seq',
                   guide=guide_legend(title='Alcohol (%)')) + 
  labs(x ='Quality (assigned level)', y= 'Volatile acidity (g/dm³)') +
  ggtitle("Relationship between alcohol percentage, volatile acidity,
          and quality level for red wines") + 
     theme(plot.title = element_text(lineheight=.8, face="bold", hjust=-0.27))
```

### Description One
This chart was suggested after my first project review and shows up the main
result of this EDA. The alcohol percentage and the volatile acidity are the
dataset attributes most related to the quality of red wines. In this chart, the
alcohol percentage was plot through buckets allowing to perceive that low 
quality wines don't have more than 12% of alcohol, while high quality wines
don't have less than 9% of alcohol. On the other hand, the volatile acidity 
is the one responsible for the vinegar unpleasant taste and one of the main 
characteristics of a bad wine. From this chart this negative correlation among
volatile acidity and quality is also clear visible. For last, I have added the
title and the units for each variable, turning this chart ready to be published.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggpairs(wines[variables], aes(alpha=0.3), 
        title= "Pairplot with the main dataset attributes")
```

### Description Two
In my opinion, this pairplot summarises the main aspects of the dataset I have 
worked with. From the univariate analysis I was able to identify the most
relevant one and then to get an overview of the data by them. The diagonal
presents the univariate distribution and upper it we have the correlation values
among the attributes. It is the indispensable starting point for any EDA.  


### Plot Three
```{r echo=FALSE, Plot_Three}
radarchart(alcohol_quality, axistype=1, calcex = 0.8, vlcex=0.9,
           caxislabels = c('9.6%','','11%','','12.4%'),
           title="Percentage of alcohol by quality level")
```

### Description Three
I confess the radarchart was the only visualization I found harder to build with
R than with python. Even so, I spent a long time trying to find the appropriate
package and finding out how to fill its parameters. The result, however, was 
exactly what I expected: a simple chart but one with the most evident 
relationships I found out in this dataset.  

------

## Reflection
Regarding the purpose of this analysis, some reflections and discussions were 
included closer to each chart during the EDA process, in order to facilitate 
their comprehension. A final word on this, though, is that the EDA allowed not
only to understand the dataset but also to identify its main characteristics, 
and two variables that are somehow correlated to the target variable, in this 
case the quality of the red wines.  

Some questions aroused from this EDA that may be explored on future 
researches:  

* What is the winery of each evaluated wine? The reason to include such 
information is to correlate the wine density with the water density and pH of 
the main hydrographic basins of the regions where the wines are produced, also 
associating these attributes with the "body" perception.  

* Using this same dataset, it would be interesting to improve the linear model 
by changing the approach to select the training data. From EDA we know the 
attributes we have chosen presen more correlation on the bordering values. In 
this case, it seems reasonably to model our linear regression taking into
account this ranges, by defining some operation thresholds for the model, for
example.  

* Also, it will be interesting to apply machine learning on this dataset, since
the EDA is already done. 

Regarding the technical aspects of EDA, I'm pretty sure R is now one more 
amazing tool I will put on my data scientist toolkit. Before I got enrolled in 
this course, and before putting my hands on dirty with this project, I thought 
someone should choose among python or R. Now I see both languages are plenty of 
resources and both can be used complementarily.


## References
<a name='[1]'></a> [1] P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  
  Available at http://dx.doi.org/10.1016/j.dss.2009.05.016
  
<a name='[2]'></a>[2] http://scitechconnect.elsevier.com/wp-content/uploads/2014/06/main-38.pdf

<a name='[3]'></a>[3] https://www.theguardian.com/lifeandstyle/2013/jun/23/wine-tasting-junk-science-analysis
